Statistical performance analysis of neural mass models?

### 2013-03-01 doc/Sigmoid_Functions.tex

@@

\section{Model}

The model takes the form
\begin{subequations}
	\label{eqn:model_general}
\begin{align}
	\vec x_k &= \vec f_k(\vec x_{k-1}) + \vec v_k \\
	\vec y_k &= \mat H\vec x_{k-1} + \vec w_k
\end{align}
\end{subequations}
where $\vec v_k$ and $\vec w_k$ are white, uncorrelated, zero mean Gaussian noise vectors with covariances of $\mat Q$ and $\mat R$, respectively. For this model the transition density is defined as
\begin{align}
	p(\vec x_k|\vec x_{k-1}) &= \mathcal N(\vec f(\vec x_{k-1})),\mat Q) \\
	&= (2\pi)^{-d/2} |\mat Q|^{-1/2} \exp\left(-\frac{1}{2}(\vec x_k - \vec f(\vec x_{k-1}))^T \mat Q^{-1} (\vec x_k - \vec f(\vec x_{k-1})) \right)
\end{align}
and the likelihood is
\begin{align}
	p(\vec y_k|\vec x_k) &= \mathcal N(\mat H\vec x_k,\mat R) \\
	&= (2\pi)^{-n/2} |\mat R|^{-1/2} \exp\left(-\frac{1}{2}(\vec y_k - \mat H\vec x_k)^T \mat R^{-1} (\vec y_k - \mat H\vec x_k) \right)
\end{align}

\section{PCRB}

The posterior Cram\'er-Rao bound (PCRB) is computed from the Fisher information matrix computed on the joint density $p(\vec y,\vec x)$,
\begin{align}
	\mat J &= \mathbb E\left[ \nabla_{\vec x}\nabla_{\vec x}^T \log p(\vec y,\vec x) \right]
\end{align}
where the expectation is taken over $\vec y$ and $\vec x$. The PCRB bounds the covariance of an unbiased estimator, $\mat P$,
\begin{align}
	\mat P &\ge \mat J^{-1}
\end{align}

Recursive formulas are based on the logarithm of the likelihood and transition densities, which for the model above are
\begin{align}
	\log p(\vec x_k|\vec x_{k-1}) &= C_1 -\frac{1}{2}(\vec x_k - \vec f(\vec x_{k-1}))^T \mat Q^{-1} (\vec x_k - \vec f(\vec x_{k-1})) \\
	\log p(\vec y_k|\vec x_k) &= C_2 -\frac{1}{2}(\vec y_k - \mat H\vec x_k)^T \mat R^{-1} (\vec y_k - \mat H\vec x_k)
\end{align}

\subsection{Fisher information recursion (non-singular $Q$)} 
A recursive formula for the FIM, $\mat J_k$, was developed in \cite{Tichavsky1998}. The recursion proceeds as
\begin{align}
	\mat J_k(\vec x_k) &= \mat W_k - \mat V_k^T\left[ \mat J_{k-1}(\vec x_{k-1}) + \mat U_k\right]^{-1}\mat V_k 
\end{align}
where
\begin{align}
	\mat U_k &= -\mathbb E\left[ \nabla_{\vec x_{k-1}}\nabla_{\vec x_{k-1}}^T \log p(\vec x_k|\vec x_{k-1}) \right] \label{eqn:recursion_U}\\
	\mat V_k &= -\mathbb E\left[ \nabla_{\vec x_{k-1}}\nabla_{\vec x_k}^T \log p(\vec x_k|\vec x_{k-1}) \right] \label{eqn:recursion_V}\\
	\mat W_k &= -\mathbb E\left[ \nabla_{\vec x_k}\nabla_{\vec x_k}^T \log p(\vec x_k|\vec x_{k-1}) \right] - \mathbb E\left[ \nabla_{\vec x_k}\nabla_{\vec x_k}^T \log p(\vec y_k|\vec x_k) \right] \label{eqn:recursion_W}
\end{align}

The derivatives for the above model are
\begin{align}
	\nabla_{\vec x_k}^T \log p(\vec x_k|\vec x_{k-1}) &= -(\vec x_k - \vec f(\vec x_{k-1}))^T\mat Q^{-1}  \\
	\nabla_{\vec x_k}\nabla_{\vec x_k}^T \log p(\vec x_k|\vec x_{k-1}) &= -\mat Q^{-1} \\
	\nabla_{\vec x_{k-1}}\nabla_{\vec x_k}^T \log p(\vec x_k|\vec x_{k-1}) &= \mat F(\vec x_{k-1})\mat Q^{-1} \\
	\nabla_{\vec x_{k-1}}^T \log p(\vec x_k|\vec x_{k-1}) &= (\vec x_k - \vec f(\vec x_{k-1}))^T\mat Q^{-1} \mat F(\vec x_{k-1}) \\
	\nabla_{\vec x_{k-1}}\nabla_{\vec x_{k-1}}^T \log p(\vec x_k|\vec x_{k-1}) &= -\mat F(\vec x_{k-1})^T\mat Q^{-1} \mat F(\vec x_{k-1}) \\
	\nabla_{\vec x_k}^T \log p(\vec y_k|\vec x_k) &= (\vec y_k - \mat H\vec x_k)^T\mat R^{-1} \mat H \\
	\nabla_{\vec x_k}\nabla_{\vec x_k}^T \log p(\vec y_k|\vec x_k) &= -\mat H^T\mat R^{-1} \mat H 
\end{align}
Substituting these into \eqref{eqn:recursion_U}--\eqref{eqn:recursion_W} gives,
\begin{align}
	 \mat U_k &= \mathbb E\left[ \mat F(\vec x_{k-1})^T\mat Q^{-1} \mat F(\vec x_{k-1}) \right ] \label{eqn:pcrb_u} \\
	\mat V_k &= \mathbb E\left[ -\mat F(\vec x_{k-1})^T\mat Q^{-1} \right ] \label{eqn:pcrb_v} \\
	\mat W_k &= \mat H^T\mat R^{-1}\mat H + \mat Q^{-1}
\end{align}

This method is not applicable when noise is only added to a subset of the states each transition. In this case, $\mat Q$ is singular and $\mat Q^{-1}$ does not exist. Another method is required.

\subsection{Cramer-Rao bound recursion (singular $Q$)}

An alternative formulation is to recursively compute the bound instead of the matrix, i.e., $\mat P_k = \mat J_k^{-1}$ directly rather than $\mat J_k$. The formulas suggested in \cite{Bergman2001} are applicable to wider class of models (e.g. multiplicative noise), denoted by a general transition function $\tilde{\vec f}(\vec x_k,\vec v_k)$. The recursion is
\begin{subequations}
\begin{align}
	\mat P_{k|k-1} &= \bar{\mat F}_k\mat P_{k-1|k-1}\bar{\mat F}_k + \bar{\mat G}_k\mat Q_k \bar{\mat G}_k^{-1} \\
		\mat P_{k|k} &= \left(\mat P_{k|k-1}^{-1} + \bar{\mat R}_{k-1}^{-1}\right)^{-1}
\end{align}
\end{subequations}
where
\begin{align}
	\bar{\mat F}_k &= \mathbb E \left[ \nabla_{\vec x_k}^T \tilde{\vec f_k}(\vec x_k,\vec v_k)\right] \\
	\bar{\mat R}_k^{-1} &= -\mathbb E\left[ \nabla_{\vec x_k}\nabla_{\vec x_k}^T \log p(\vec y_k|\vec x_k) \right] \\
	\bar{\mat G}_k &= \mathbb E \left[ \nabla_{\vec v_k}^T \tilde{\vec f_k}(\vec x_k,\vec v_k)\right] \\
	\mat Q_k^{-1} &= -\mathbb E \left[ \nabla_{\vec v_k}\nabla_{\vec v_k}^T \log p(\vec v_k) \right]
\end{align}
and $\mat P_0^{-1}$ is the inverse of the prior covariance. The model in \eqref{eqn:model_general} is represented by $\tilde{\vec f}(\vec x_k,\vec v_k) = \vec f(\vec x_k) + \vec v_k$. This simplifies the derivatives to
\begin{align}
	\nabla_{\vec x_k}^T \tilde{\vec f_k}(\vec x_k,\vec v_k) &= \mat F_k(\vec x_k) \\
	\nabla_{\vec v_k}^T \tilde{\vec f_k}(\vec x_k,\vec v_k) &= \mat I_{d\times d} \\
	\nabla_{\vec x_k}\nabla_{\vec x_k}^T \log p(\vec y_k|\vec x_k) &= -\mat H^T\mat R^{-1} \mat H 
\end{align}
and the corresponding matrices to	
\begin{align}
	\bar{\mat F}_k &= \mathbb E\left[ \mat F_k(\vec x_k)\right] \label{eqn:pcrb_term_F}\\
	\bar{\mat R}_k^{-1} &= \mat H^T\mat R^{-1} \mat H \\
	\bar{\mat G}_k &= \mat I_{d\times d} 
\end{align}
Notice that $\mat Q^{-1}$ is not directly required in the recursion, so this method is applicable to neural field models where process noise is only added to a single state.

\subsection{Monte Carlo approximations}

The expectations in \eqref{eqn:pcrb_u} and \eqref{eqn:pcrb_v} or \eqref{eqn:pcrb_term_F} can be approximated by numerical integration using many trajectory realisations. That is,
\begin{align}
	\mat U_k &\approx \frac{1}{M}\sum_{i=1}^M \mat F(\vec x_{k-1}^{(i)})^T\mat Q^{-1} \mat F(\vec x_{k-1}^{(i)}) \\
	\mat V_k &\approx \frac{1}{M}\sum_{i=1}^M \mat F(\vec x_{k-1}^{(i)})^T\mat Q^{-1}  
\end{align}
or alternatively,
\begin{align}
	\bar{\mat F}_{k-1} &\approx \frac{1}{M}\sum_{i=1}^M \mat F(\vec x_{k-1}^{(i)})
\end{align}

\subsection{Analytical evaluation of expectations}

For the JR model the expectation in \eqref{eqn:pcrb_term_F} above is
\begin{align}
	\bar{\mat F}_k &= \mathbb E\left[ \mat F_k(\vec x_k)\right] \\
	&= \mat F_k^{\rm (lin)} + \mathbb E\left[ \mat G_k(\vec x_k)\right] 
\end{align}
where $\mat G_k(\vec x_k)$ contains the nonlinear component of the Jacobian. These nonlinear terms are derivatives of the error function and have the form,
\begin{align}
	\exp(-(x-v_0)^2/\zeta^2)
\end{align}
The expectation of such terms can only be evaluated when we know the prior mean and covariance at all times.

\section{Pendulum example}

Let $\vec x_k = [a_k\quad b_k]^T$ where $a_k$ is the pendulum angle and $b_k$ is the angular velocity,
\begin{align}
	\vec f(\vec x_{k}) &= \left[\begin{array}{c}
								a_k + b_k\Delta \\ -b_k - g\sin(a_k)\Delta
								\end{array}\right]
\end{align}
The linearisation used for the EKF is
\begin{align}
	\mat F(\vec x_{k}) &= \left[\begin{array}{cc}
					1 & \Delta \\
					-g\cos(a_k) & 1 \\
				\end{array}\right]
\end{align}
The observation matrix is $H=[1\quad 0]$, so that only the first state element is measured.

\subsection{Simulations}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\textwidth]{../figures/pendulum_trajectory}
	\caption{Trajectory of a pendulum and EKF estimates}
\end{figure}
\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\textwidth]{../figures/pendulum_pcrb}
	\caption{MSE of EKF estimates (averaged over 1000 realisations) compared to the PCRB. The suboptimal EKF is slightly above the PCRB.}
\end{figure}

\small
\bibliographystyle{ieeetr}
\bibliography{references}
% \bibliography{../../library}

>>>>

\DeclareMathOperator*{\erf}{erf}

@@

\section{Sigmoid functions}

The sigmoid can be specified using the logistic function
\begin{align}
	g(x) = \frac{1}{1 + \exp(\varsigma(v_0 - x))}
\end{align}
which has the derivative
\begin{align}
	g'(x) = \varsigma g(x)(1 - g(x))
\end{align}

Alternatively, using the error function
\begin{align}
	g(x) = \frac{1}{2}\left( 1 + \erf\left( \frac{x - v_0}{\xi\sqrt{2}}\right)\right)
\end{align}
which has the derivative
\begin{align}
	g'(x) = \frac{1}{\xi\sqrt{2\pi}} \exp\left(-\left(\frac{x - v_0}{\xi\sqrt{2}}\right)^2 \right) 
\end{align}

\section{Relationship between widths}

Through numerical simulation we can deduce that setting the error function width to
\begin{align}
	\xi = \frac{1.699}{\varsigma}
\end{align}
gives a sigmoid with approximately the same shape as a logistic function with width $\varsigma$. This relationship seems to hold across a wide range of widths. It is likely this number could be derived analytically.

% \bibliography{../../library}

### 2013-03-01 manuscript/manuscript.tex

\DeclareMathOperator*{\erf}{erf}

\title{PCRB}

\date{\today}

@@

\section{Sigmoid functions}

The sigmoid can be specified using the logistic function
\begin{align}
	g(x) = \frac{1}{1 + \exp(\varsigma(v_0 - x))}
\end{align}
which has the derivative
\begin{align}
	g'(x) = \varsigma g(x)(1 - g(x))
\end{align}

Alternatively, using the error function
\begin{align}
	g(x) = \frac{1}{2}\left( 1 + \erf\left( \frac{x - v_0}{\xi\sqrt{2}}\right)\right)
\end{align}
which has the derivative
\begin{align}
	g'(x) = \frac{1}{\xi\sqrt{2\pi}} \exp\left(-\left(\frac{x - v_0}{\xi\sqrt{2}}\right)^2 \right) 
\end{align}

\section{Relationship between widths}

Through numerical simulation we can deduce that setting the error function width to
\begin{align}
	\xi = \frac{1.699}{\varsigma}
\end{align}
gives a sigmoid with approximately the same shape as a logistic function with width $\varsigma$. This relationship seems to hold across a wide range of widths. It is likely this number could be derived analytically.

% \bibliography{../../library}

>>>>

\title{Statistical performance analysis of neural mass models?}
\date{\today}

@@

\section{Introduction}

History of models and estimation attempts

Usefulness of bound: benchmark algorithms, experimental design, insight about inherent uncertainty -- which do we address

\section{Background}

Present 3 models?

\section{Theory / Performance Bounds}

Introduce recursive computation of bound and Monte Carlo approximation

\section{Methods?}

\section{Results}

\subsection{Effect of increasing model complexity -- bound should increase}

e.g. Bound for model A > model B > model C

\subsection{Effect of parameters changing -- bound should change}

e.g. seizure easier to estimate than alpha rhythms

\subsection{Effect of measurement noise? -- pretty boring}

e.g. scalp EEG vs iEEG?

\subsection{Augmented models (joint estimation of parameters and states)?}

\section{Discussion}

extensions to other applications e.g. experimental design

contextualize work w.r.t. existing literature?

\section{Conclusion}

% \small
% \bibliographystyle{ieeetr}
% \bibliography{references}
